---
title: "march_report"
format: pdf
---

```{r}
# Load necessary libraries
library(dplyr)
library(tidyr)
install.packages('tinytex')
tinytex::install_tinytex()


events_full <- read.csv("/Users/daraun/Downloads/2025_02_18_utilization_events.csv")
reports_full <- read.csv("/Users/daraun/Downloads/2025_02_18_utilization_report.csv")

head(events_full)
head(reports_full)
```

# Creating a table that is filtered by different variables to achieve the outcome wanted.

```{r}
pre_ibis_inpatient_table <- events_full %>% filter(org_name == "Unicare - Study") %>% filter(pre_ibis_inpatient == 1 | event_type == "no_utilization") %>%
  group_by(pid, event_type) %>%
  summarize(count = n(), .groups = "drop") %>% pivot_wider(names_from = event_type, values_from = count) %>% replace(is.na(.), 0) %>% rename(ibis_inpatient_count = inpatient)
```

Retrieving the counts from the newly created table.

```{r}
pre_ibis_inpatient_table %>% count(ibis_inpatient_count)
```

# Getting the counts from each period (pre/during/post-IBIS).

```{r}
# total admissions
pre_ibis_data <- events_full %>% filter(org_name == "Unicare - Study" & pre_ibis_inpatient == 1)
during_ibis_data <- events_full %>% filter(ibis_inpatient == 1 & org_name == "Unicare - Study")
post_ibis_data <- events_full %>% filter(pre_ibis_inpatient != 1 & ibis_inpatient == 0 & org_name == "Unicare - Study")

# count totals for each period
total_admissions_pre_ibis <- nrow(pre_ibis_data)
total_admissions_during_ibis <- nrow(during_ibis_data)
total_admissions_post_ibis <- nrow(post_ibis_data)

# print the counts
print(total_admissions_pre_ibis)
print(total_admissions_during_ibis)
print(total_admissions_post_ibis)
```

# Calculating total patient years for each period 

```{r}
# calculate total patient years for each period by summarizing total covered days per patient
pre_ibis_patient_data <- pre_ibis_data %>% group_by(pid) %>%
  summarise(total_days_pre_ibis = sum(covered_days_on_ibis, na.rm = TRUE))

during_ibis_patient_data <- during_ibis_data %>% group_by(pid) %>%
  summarise(total_days_during_ibis = sum(covered_days_on_ibis, na.rm = TRUE))

# convert covered days into patient years
pre_ibis_patient_data <- pre_ibis_patient_data %>%
  mutate(patient_years_pre_ibis = total_days_pre_ibis / 365)

during_ibis_patient_data <- during_ibis_patient_data %>%
  mutate(patient_years_during_ibis = total_days_during_ibis / 365)

# calculate total patient years for each period
total_patient_years_pre_ibis <- sum(pre_ibis_patient_data$patient_years_pre_ibis, na.rm = TRUE)
total_patient_years_during_ibis <- sum(during_ibis_patient_data$patient_years_during_ibis, na.rm = TRUE)

# per 1000 patient years
admissions_per_1000_pre_ibis <- (total_admissions_pre_ibis / total_patient_years_pre_ibis) * 1000
admissions_per_1000_during_ibis <- (total_admissions_during_ibis / total_patient_years_during_ibis) * 1000

# results
print(admissions_per_1000_pre_ibis)
print(admissions_per_1000_during_ibis)
```

# Calculating total patient days for each period.

```{r}
# calculate total inpatient days for each period
pre_ibis_inpatient_data <- pre_ibis_data %>% group_by(pid) %>%
  summarise(total_inpatient_days_pre_ibis = sum(pre_ibis_inpatient_days, na.rm = TRUE))

during_ibis_inpatient_data <- during_ibis_data %>% group_by(pid) %>%
  summarise(total_inpatient_days_during_ibis = sum(ibis_inpatient_days, na.rm = TRUE))

# calculate total inpatient days for each period
total_inpatient_days_pre_ibis <- sum(pre_ibis_inpatient_data$total_inpatient_days_pre_ibis, na.rm = TRUE)
total_inpatient_days_during_ibis <- sum(during_ibis_inpatient_data$total_inpatient_days_during_ibis, na.rm = TRUE)

#per 1000 patient days
inpatient_days_per_1000_pre_ibis <- (total_inpatient_days_pre_ibis / total_patient_years_pre_ibis) * 1000
inpatient_days_per_1000_during_ibis <- (total_inpatient_days_during_ibis / total_patient_years_during_ibis) * 1000

# print the results
print(inpatient_days_per_1000_pre_ibis)
print(inpatient_days_per_1000_during_ibis)

```

# This code computes the admit counts for patients pre/during/post-IBIS.

```{r}
# count number of admissions per patient for pre-Ibis
pre_ibis_counts <- events_full %>%
  filter(pre_ibis_inpatient == 1, event_type == "inpatient") %>%
  group_by(pid) %>%
  summarise(admit_count = n()) %>%
  count(admit_count)

# count number of admissions per patient for during-Ibis
during_ibis_counts <- events_full %>%
  filter(ibis_inpatient == 1, event_type == "inpatient") %>%
  group_by(pid) %>%
  summarise(admit_count = n()) %>%
  count(admit_count)

post_ibis_counts <- events_full %>%
  filter(ibis_inpatient == 0, event_type == "inpatient") %>%
  group_by(pid) %>%
  summarise(admit_count = n()) %>%
  count(admit_count)

# display results
pre_ibis_counts
during_ibis_counts
post_ibis_counts
```

# By taking the total number of covered days per patient, this code is used to calculate covered years. 

```{r}
# calculate total patient years for each period by summarizing total covered days per patient
pre_ibis_patient_data <- pre_ibis_data %>% group_by(pid) %>%
  summarise(total_days_pre_ibis = sum(covered_days_on_ibis, na.rm = TRUE))


during_ibis_patient_data <- during_ibis_data %>% group_by(pid) %>%
  summarise(total_days_during_ibis = sum(covered_days_on_ibis, na.rm = TRUE))

pre_ibis_patient_data
during_ibis_patient_data
```

# Wilcoxon Test

```{r}
# Run the Wilcoxon signed-rank test
wilcoxon_result <- wilcox.test(total_admissions_pre_ibis, total_admissions_post_ibis, paired = TRUE)
wilcoxon_result
```

# Breaking down the results:

V = 0: This means that the sum of the positive ranks is zero, indicating that all differences between paired observations are either zero or negative.

p-value = 1: A p-value of 1 suggests that there is no evidence to reject the null hypothesis. In other words, the observed data is perfectly consistent with the null hypothesis.

Alternative hypothesis: The test was conducted to determine if the true location shift (i.e., the median difference between the two groups) is not equal to 0.

Interpretation: Since the p-value is very high, it suggests that there is no significant difference between total_admissions_pre_ibis and total_admissions_post_ibis. The intervention (IBIS) does not appear to have changed the total admissions in a statistically significant way.

# Log Model

```{r}
logit_model <- glm(ibis_inpatient ~ pre_ibis_inpatient, data = events_full, family = "binomial")

summary(logit_model)
```

# Results & its meaning:

Intercept: this represents the log-odds of being an ibis inpatient when pre_ibis_inpatient = 0.

-16.76, p = 0.959 shows that there is no relationship between being a pre-ibis inpatient and ibis inpatient. As well as the p-value being very high showing that this effect is not statistically significant. pre_ibis_inpatient does not influence ibis_inpatient.

z-value being close to zero means that it is not statistically significant.

```{r}
# McNemar’s Test: Check probability of admission pre vs. post.
pre_post_matrix <- matrix(c(
  sum(pre_ibis_counts$admit_count == 0 & during_ibis_counts$admit_count > 0),
  sum(pre_ibis_counts$admit_count > 0 & during_ibis_counts$admit_count == 0),
  sum(pre_ibis_counts$admit_count > 0 & during_ibis_counts$admit_count > 0),
  sum(pre_ibis_counts$admit_count == 0 & during_ibis_counts$admit_count == 0)
), nrow = 2, byrow = TRUE)

mcnemar_result <- mcnemar.test(pre_post_matrix)
mcnemar_result
```

# Results Broken Down:

McNemar's chi-squared = 12.071:

This test statistic is based on the expected values in the paired categories. A higher value suggests a stronger difference between the two conditions (pre & post).

df = 1 due to the test taking in only two conditions (pre & post).

The p-value of 0.000512 shows that there is strong evidence against the null hypothesis, meaning that there is a strong statistically significant between the pre & post conditions.
